---
title: "ANOVA Weeks 1 - 3"
author: Chris Ives
output: 
  xaringan::moon_reader:
    css: ["default", "new.css"]
    lib_dir: libs
    nature:
      navigation:
        scroll: false
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
needs(tidyverse, xaringan, rio, here, knitr, kableExtra, misty)
data <- read.sav(here("data/Lab2_Vocab.sav"), use.value.labels = T)
```

# Contents

* Foundations
  + Importing Data
  + Descriptives
  + Plots
* ANOVA
  + One-way ANOVA
  + Pairwise Comparisons
  + Planned Contrasts
  + Power Analysis
* Appendix
  + Custom Contrast Sets

---
## Installing the `needs` Package

- Can install packages and load them
- No longer need `install.packages()` and `library()`
- It will ask if you would like to load it every time RStudio opens. Select yes.

```{r, eval = FALSE}
install.packages("needs")
```

---
## Importing the Data
Normally we would use the `import()` function to import our data. However, with the SPSS `.sav` files we don't get factor labels.
``` {r}
data_import <- import(here("data/Lab2_Vocab.sav"))
head(data_import)
```

Notice how we have 1s down the instruct column.

---
## The Fix
For now, when working with `.sav` files we will use the `read.sav` function from the `misty` package. `use.value.labels = TRUE` tells it to use the labels as the cell values.

``` {r}
needs(misty)
l2_data <- misty::read.sav(here("data/Lab2_Vocab.sav"), 
                        use.value.labels = TRUE)
head(l2_data)
```

---
## Checking The Assumptions of ANOVA

- Independence of observations
- Normality
- Homogeneity of variance

---
## Descriptive Statistics

Make sure the `psych` package is installed and loaded:
```{r, echo = TRUE}
needs(psych)
```

`describe` from the `psych` package is one of the more popular functions for descriptive statistics.

```{r, echo = TRUE, eval = FALSE}
describe(l2_data)
```

```{r echo = FALSE, eval = TRUE}
d <- describe(l2_data)
rmarkdown::paged_table(d)
```

---
## Independence of Observations
 - Do participants cross groups?
 - We know this if we know the study design
 - In this study, each student experienced one and only one lecture

---
## Normality
- We can check the skew and kurtosis values from our `describe()` output.
- 0 ± 2 is a good rule of thumb for a tenable assumption of normality.

```{r echo = FALSE, eval = TRUE}
d <- as.data.frame(d) %>% 
  select(vars, n, skew, kurtosis)
rmarkdown::paged_table(d)
```

- Instruct is a categorical variable, so we can ignore the skew and kurtosis on that. 

---
## Visual Inspection for Normality
- For the most part, plots will be wrapped using the `ggplot()` function.

### Histogram
```{r, echo = TRUE, eval = FALSE}
ggplot(data = l2_data, aes(x = vocab)) +
  geom_histogram()
```

- `aes()` refers to aesthetics. What are the variables we want represented in our plots? Since we just want counts of a single continuous variable, we just need to specify our `x` (i.e., `x = vocab`).

---
## Histograms
```{r, echo = TRUE, eval = TRUE, message = F, fig.width = 8, fig.height= 6, fig.align='center'}
ggplot(data = l2_data, aes(x = vocab)) +
  geom_histogram()
```

---
## Boxplot
- To get boxplots, we just substitute `geom_histogram()` for `geom_boxplot()` and modify our aesthetics.

```{r, echo = TRUE, fig.width = 6, fig.height= 4, fig.align='center'}
ggplot(data = l2_data, aes(x = instruct, y = vocab)) +
  geom_boxplot()
```

---
## Homogeneity of Variance
Homogeneity test is a separate analysis. We'll just use the `leveneTest` function for the `car` package. Formula is the same as for the ANOVA we want to run. Specify `center = "mean"` (function's default is median) to match SPSS results.

``` {r}
needs(car)
car::leveneTest(vocab ~ instruct, data = l2_data, center = "mean")
```

Our significant result shows error variance around the *mean* is not equal across groups.
---
class: inverse-blue middle
# Running ANOVAs

---

Example here is from Lab 1 data.
```{r, echo = F}
l1_data <- misty::read.sav(here("data/Lab1_Vocab.sav"), 
                        use.value.labels = TRUE)
```

ANOVAs will be run using the `anova_test()` function. Formula is specified as `DV ~ IV`. My convention is to number my model objects (e.g., m1, m2, etc.)

```{r, messages = FALSE}
needs(rstatix)
m1 <- anova_test(data = l1_data, formula = vocab ~ instruct, detailed = TRUE)
m1

```

---
## Effect Sizes
Effect sizes can be specified using `effect.size = ____` in the `anova.test()` function. Use `effect.size = "ges"` for generalized eta squared or `effect.size = "pes"` for partial eta squared. Default is "`ges"`.

```{r, messages = FALSE, eval = F}
anova_test(data = l1_data, formula = vocab ~ instruct,
           effect.size = "ges")
```

---

```{r, messages = FALSE, eval = T}
anova_test(data = l1_data, formula = vocab ~ instruct,
           effect.size = "ges")
```

```{r}
anova_test(data = l1_data, formula = vocab ~ instruct,
           effect.size = "pes")
```

Notice they are the same for a one-way between subjects ANOVA.
---
## Posthoc Comparisons | Tukey
If you write out `"data = l2data"` the function will produce an error since it designed for multiple types of objects. Here we just type "l2data".
```{r, eval = F}
rstatix::tukey_hsd(l2_data, formula = vocab ~ instruct)
```

```{r, echo = F}
tukey <- rstatix::tukey_hsd(l2_data, formula = vocab ~ instruct)

rmarkdown::paged_table(tukey)
```

---
## Posthoc Comparisons | Games-Howell
Remember, Games-Howell is used if the assumption of variance homogeneity is violated.
```{r, eval = F}
rstatix::games_howell_test(vocab ~ instruct, data = l2_data, 
                           conf.level = 0.95, 
                           detailed = FALSE)
```

```{r, echo = F}
gh <- rstatix::games_howell_test(vocab ~ instruct, data = l2_data, 
                                 conf.level = 0.95, 
                                 detailed = FALSE)
rmarkdown::paged_table(gh)
```

---
## Posthoc Comparisons | Bonferroni
``` {r, eval = F}
pairwise_t_test(data = l2_data, vocab ~ instruct, 
                p.adjust.method = "bonferroni")
```

``` {r, echo = FALSE}
bf <- pairwise_t_test(data = l2_data, vocab ~ instruct, 
                p.adjust.method = "bonferroni")

rmarkdown::paged_table(bf)
```

---
class: inverse-green middle
# Specified Contrasts
## (Lab 3)
* They are not automatic in R and are a bit difficult. It is not important that you understand the reasoning behind all of the following steps.
---
## Specified Contrasts
``` {r, echo = F}
l3_data <- misty::read.sav(here("data/Lab3_Vocab.sav"),
                        use.value.labels = TRUE)
```

Check the order of your levels.
```{r}
levels(l3_data$instruct)
```

Make sure they are in the order that you would like and code your contrasts accordingly. As an example, If they need to be reordered use:
``` {r}
l3_data$instruct <- factor(l3_data$instruct, levels = c("social science", "history", "physical science"))
```
Note: We will be using the original order in the following contrasts.

``` {r, echo = F}
l3_data$instruct <- factor(l3_data$instruct, levels = c("physical science", "social science", "history"))

```

---
**Step 1**: Set your contrasts. We'll do Helmert here.
``` {r, eval = T}
contrast1 <- c(1, -.5, -.5) # physical science vs others
contrast2 <- c(0, 1, -1) #social science vs history
```

**Step 2**: Bind the vectors into a temporary matrix. Constant should be equal to 1/(length of your vectors).
```{r}
mat.temp <- rbind(constant=1/3, contrast1, contrast2)
mat.temp
```

---
**Step 3**: Take the inverse of the matrix using the `solve()` function. Then we are dropping the first column with the constants.
```{r}
mat <- solve(mat.temp)
mat
mat<- mat[ , -1]
mat
```

---
**Step 4**: Run your model formula using `lm()` and set contrasts. Here we are linking our "instruct" variable with our contrast matrix. Remember:
* contrast1 = physical science vs others
* contrast2 = social science vs history

``` {r}
m_contrasts <- lm(vocab ~ instruct, data=l3_data, contrasts = list(instruct = mat))
summary(m_contrasts)
```

---
## How do I get the SDs for calculating effect sizes?

We will use the `mutate` function to code new groupings. Then, we will run descriptives on this variable. My `case_when` function reads: If `instruct` = `social science` or `instruct` = `history`, code contrast1 as `humanities`. If `instruct` = `physical science` code as `physical science`.

``` {r}
l3_data <- l3_data %>% 
  mutate(contrast1 = case_when(instruct == "social science" | instruct == "history" ~ "humanities",
                               instruct == "physical science" ~ "physical science"))

contrast1_desc <- describe(l3_data ~ contrast1)
```

---

```{r}
rmarkdown::paged_table(contrast1_desc$humanities)
```

```{r}
rmarkdown::paged_table(contrast1_desc$`physical science`)
```

---

## Pooling SDs

Here I am saving the second element of each descriptive table (since vocab is the second row in the data frame) to calculate pooled SDs.
``` {r}
sd1 <- contrast1_desc$humanities$sd[2]
sd2 <- contrast1_desc$`physical science`$sd[2]

cohens_sd <- sqrt((sd1^2 + sd2^2)/2)
cohens_sd

n1 <- contrast1_desc$humanities$n[2]
n2 <- contrast1_desc$`physical science`$n[2]

hedges_sd <- sqrt(((n1-1)*sd1^2 + (n2-1)*sd2^2)/(n1+n2-2))
hedges_sd
  
```

---
## Calculate Hedges' g

Now that we have our pooled SD and coefficient, we just need to run the calculation for contrast1.

```{r}
coefficients(m_contrasts)
```

```{r}
hedges_g <- 9.75/hedges_sd
hedges_g
```

---
## Calculate Effect Size

For an unbalanced dataset, we will use `pwr.2p2n.test` from the `pwr` package to run a power analysis on our contrast. `h` will be equal to our effect size. We already saved our n's earlier, so n1 will just be equal to n1 (same for n2). 

```{r}
needs(pwr)

pwr.2p2n.test(h = hedges_g, n1 = n1, n2 = n2, power = NULL, sig.level = .05)
```

---
## What about my ANOVA results?
To get the anova output, just run `rstatix::anova_test()` on your `m_contrasts` object. SPSS uses Type III SS by default, so I am matching it here.
```{r}
contrast_anova <- anova_test(m_contrasts, 
                 type="III",
                 effect.size = "ges",
                 detailed = TRUE)
contrast_anova
```
---
class: inverse-red middle
# Appendix

---

## Contrasts

**Isolated vs. Orthogonal Sets**.
If you are conducting an isolated contrast, or non-orthogonal set, you will need to use a different process for specifying contrasts. Here I demonstrate doing a single contrast. The primary difference is using `ginv()` from the `MASS` package rather than `solve()`.

```{r}
contrast0 <- c(1, -.5, -.5) 
mat.temp <- rbind(constant = 1/2, contrast0)
mat.temp
mat <- MASS::ginv(mat.temp)
mat <- mat[ , -1]
mat
```

---

Essentially, R fills in the rest of the matrix. This means that there will be "extra" contrasts in the output that you will need to ignore. Here `instruct1` corresponds to my first and only contrast. `instruct2` can be ignored. 
```{r}
m_contrasts <- lm(vocab ~ instruct, data=l3_data, contrasts = list(instruct = mat))
summary(m_contrasts)
```
---

If you want to make it easier to refer back to your contrasts, you can rename them in your output using the following code. Here I chose to keep the name of the intercept, name my contrast, and make everything else unnamed.

```{r}
attr(m_contrasts$coefficients, "names") <- c("Intercept", "Phys Sci vs. Others", "")
summary(m_contrasts)
```