---
title: "EDUC 640"
subtitle: "Two-Way Factorial ANOVA"
author: Chris Ives
output: 
  xaringan::moon_reader:
    css: ["default", "new.css"]
    lib_dir: libs
    nature:
      navigation:
        scroll: false
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
needs(tidyverse, xaringan, rio, here, misty, emmeans, rstatix)
lb5 <- read.sav(here("data/Lab5_Vocab.sav"), use.value.labels = T)
```

# Contents

* Plotting Means
* Recoding/Reordering Factors
* Contrasts using `emmeans`
---

## Plot Marginal Means

Here I am creating a new summary data frame. After grouping by `meth` and `inst`, I just make `vocab` equal to the mean of those groups. This dataframe is another way of getting marginal mean values.

```{r}
means <- lb5 %>% 
  group_by(inst, meth) %>% 
  summarise(vocab = mean(vocab))
```

---

```{r}
ggplot(means, aes(x = inst, y = vocab, group = meth)) +
  geom_point() +
  geom_line(aes(color = meth))
```

---
## Contrasts using `emmeans`

To do two-factor contrasts we can use the `emmeans` package. You will probably find this easier than the previous approaches where I had you specify your coding in a matrix. Whether you prefer this for one-way contrasts will be up to you.
---


## Two Factor Groupings

First we run our model specification using `lm`. Then we run `emmeans` on that object to calculate the marginal means. There are multiple ways to use `emmeans` if you plan to run follow-up pairwise comparisons. This first way will run comparisons within the `by` grouping.
```{r}
model <- lm(vocab ~ meth + inst + meth:inst, lb5)
means_1 <- emmeans(model, "inst", by = "meth")
means_1
```
---
```{r}
pairs(means_1)
```
---

This way doesn't "nest" one factor in the other and will calculate all possible comparisons.

```{r}
means_2 <- emmeans(model, ~inst*meth)
means_2
```
---

```{r, eval = F}
pairs(means_2)

```

```{r, echo = F}
pairs <- as.data.frame(pairs(means_2))
rmarkdown::paged_table(
  pairs, options = list(rownames.print = F)
)
```

---

## Specifying contrasts

We'll be using the listing structure from the second method. We just have to specify our contrasts into a list.
```{r}
means_2
```

```{r}
contrasts <- list(
  hyp1 = c(1, -.5, -.5, -1, .5, .5),
  hyp2 = c(1, -1, 0, -1, 1, 0)
)
```

---

Much easier!!

```{r}
contrast(means_2, contrasts, adjust = "holm")
```

---

## Effect Sizes

If you want partial eta effect sizes (like SPSS output), you'll need to calculate them from the contrast output using the function below.

```{r}
effectsize::t_to_eta2(
  t = c(-1.043, -2.628),
  df_error = 30
)
```

---

# Appendix
All of the following slides reflect less efficient ways of doing contrasts. I am just leaving them in if you want to see some examples of renaming levels and uniting variables.

---
## Recoding Data

We have to start by combining our two factors into one factor with six levels. The level names are a little long for my taste so here I am shortening them. First check the orders of the levels so your names are assigned appropriately.
``` {r, messages = F}
levels(lb5$meth)
levels(lb5$meth) <- c("comp", "stan")
```

```{r}
levels(lb5$inst)
levels(lb5$inst) <- c("phys", "soc", "hist")
```
---

Next I join the two factors into one with `unite`. `col =` specifies the name of the new column, followed by the columns I am joining. I chose to separate them with "_" and set `remove = FALSE` so I don't delete the old variables.

```{r}
lb5 <- lb5 %>% 
  unite(col = ivs, meth, inst, sep = "_", remove = FALSE)

head(lb5)
```
---

Last bit of data prep is order them in a way that will make sense for me when I code out contrasts later. This order reflects what's in Gina's slides.

```{r}
lb5$ivs <- ordered(lb5$ivs, c("comp_phys", "comp_soc", "comp_hist", "stan_phys", "stan_soc", "stan_hist"))

levels(lb5$ivs)
```

---

## Emmeans
Start by specifying your model and then running emmeans on that model. Note that I am using `ivs`, which is our combined two factor variable.
```{r}
m1 <- lm(vocab ~ ivs, data = lb5)
emm <- emmeans(m1, ~ ivs)
```

---

Then we will check the level order and assign a vectors to each. We have 6 levels so the vector is 6 numbers long. A `1` in the vector means I am saving the mean of that level to my object. So, the mean of "comp_phys" is saved to `A1B1`.

```{r}
levels(lb5$ivs)
A1B1 <- c(1, 0, 0, 0, 0, 0)
A1B2 <- c(0, 1, 0, 0, 0, 0)
A1B3 <- c(0, 0, 1, 0, 0, 0)
A2B1 <- c(0, 0, 0, 1, 0, 0)
A2B2 <- c(0, 0, 0, 0, 1, 0)
A2B3 <- c(0, 0, 0, 0, 0, 1)
```
---

## Hypothesis 1

Now we can specify our contrasts (reference slides 445-446 in Lab 5). This runs the contrasts so it prints the coding scheme.

```{r}
contrast(emm, method = list(
  (A1B1 - (A1B2 + A1B3)/2) -
  (A2B1 - (A2B2 + A2B3)/2)
  ))
```
Here I name the contrast.
```{r}
contrast(emm, method = list(
  "Hyp1" = (A1B1 - (A1B2 + A1B3)/2) -
  (A2B1 - (A2B2 + A2B3)/2)
))
```
---
# Hypothesis 2

Same process for our second contrast.
```{r}
contrast(emm, method = list(
  (A1B1 - A1B2) - (A2B1 - A2B2)
  ))
```

```{r}
contrast(emm, method = list(
  "Hyp2" = (A1B1 - A1B2) - (A2B1 - A2B2)
  ))
```
---
## Combining Contrasts

If I was planning to put them into a markdown document, I'd probably want to write it all out with one command and output.
```{r}
contrast(emm, method = list(
  "Hyp1" = (A1B1 - (A1B2 + A1B3)/2) -
  (A2B1 - (A2B2 + A2B3)/2),
  "Hyp2" = (A1B1 - A1B2) - (A2B1 - A2B2)
))

```

---

Here's the process using the method I described in the Appendix of Wk1-3 slides.

```{r}
contrast0 <- c(1, -.5, -.5, -1, .5, .5) 
mat.temp <- rbind(constant = 1/6, contrast0)
mat.temp
mat <- MASS::ginv(mat.temp)
mat <- mat[ , -1]
mat

```

---

Remember I only specified one contrast so we only pay attention to the first coefficient (that's not the intercept).
```{r}
m_contrasts <- lm(vocab ~ ivs, data=lb5, contrasts = list(ivs = mat))
summary(m_contrasts)
```

---

