---
title: "EDUC 640"
subtitle: "Within Subjects ANOVA"
author: Chris Ives
output: 
  xaringan::moon_reader:
    css: ["default", "new.css"]
    lib_dir: libs
    nature:
      navigation:
        scroll: false
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
needs(tidyverse, xaringan, rio, here, misty, emmeans, rstatix)
lb6 <- read.sav(here("data/Lab6_Vocab.sav"), use.value.labels = T)
```

# Contents

* Restructuring Data
* Within-Subjects ANOVA
* Pairwise Comparisons
* Appendix
    + Contrasts
---

## Restructuring Data

We need to restructure the data into a "long" format for ANOVA in R. I start by renaming the variables and then use `pivot_longer` to pivot those variables into two columns. The names of the columns become `inst` and the values become `vocab`.

```{r}
lb6 <- lb6 %>% 
  rename(phys = vocab1,
         social = vocab2,
         hist = vocab3) %>% 
  pivot_longer(cols = c(phys, social, hist), values_to = "vocab", names_to = "inst")
```
---

Then we can just make a normal boxplot.
```{r}

ggplot(lb6, aes(x = inst, y = vocab)) +
  geom_boxplot()
```

---
## Within Subjects ANOVA
`aov_car` will allow us to run `emmeans` on the object and uses the same method for pairwise comparisons as SPSS. We just need to specify our ID variable and our IVs within `Error()`. Mauchly's test is automatically included in the output.
```{r}
needs(afex)
m1 <- afex::aov_car(vocab ~ inst + Error(idnum/inst),
                           data=lb6, include_aov = F)

```

---
```{r}
summary(m1)
```

---

Don't worry about within-subjects contrasts. We can ignore it and don't have to worry about it. Gina says so :)

However, a brief demonstration is included in the Appendix.

---

## Pairwise Comparisons
Pairwise comparsions are similar to how we did contrasts last week. We run `emmeans` on our model using our formula (`~inst`) and then run pairs on that object.

```{r}
em <- emmeans(m1, ~inst)
pairs(em)
```
---

`pairs` does not automatically generate confidence intervals. If you needed them, you can wrap it with `confint` or run it on your pairs object.

```{r}
confint(pairs(em))


```
---
class: inverse-blue middle
# Appendix
---

## Contrasts

Contrasts require just a different step to our pairwise comparions. We just create our contrasts in a list and run `contrast` on our emmeans object. If it was a 2-way Within-Subjects ANOVA, we would have to run it on our pairs object.

```{r}
contrasts <- list(
  hyp1 = c(-.5, 1, -.5),
  hyp2 = c(0, 1, -1)
)

```

---

```{r}
contrast(em, contrasts)
```
