---
title: "EDUC 640"
subtitle: "Two-Way Within Subjects ANOVA"
author: Chris Ives
output: 
  xaringan::moon_reader:
    css: ["default", "new.css"]
    lib_dir: libs
    nature:
      navigation:
        scroll: false
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
needs(tidyverse, xaringan, rio, here, misty, emmeans, rstatix, effectsize)
lb7 <- read.sav(here("data/Lab7_Rutherford.sav"), use.value.labels = T)
```

# Contents

* Restructuring Data
* Two-Way Within-Subjects ANOVA
* Pairwise Comparisons
* Contrasts
---

## Restructuring Data (Review)

We need to restructure the data into a "long" format for ANOVA in R. I start by renaming the variables and then use `pivot_longer` to pivot those variables into two columns. The names of the columns become `condition` and the values become `total`.

```{r}
lb7 <- lb7 %>%
  rename(mem_30 = A1B1,
         mem_60 = A1B2,
         mem_180 = A1B3,
         mne_30 = A2B1,
         mne_60 = A2B2,
         mne_180= A2B3) %>%
  pivot_longer(cols = c(mem_30, mem_60, mem_180, mne_30, mne_60, mne_180), values_to = "total", names_to = "condition")
```
---
## Added Step
To split our factor into its two groupings, we need to use the `separate` function. I am specifiying that condition be separated into `condition` and `study_time`, which can be differentiated by `_`. 

```{r}
lb7 <- lb7 %>% 
  separate(condition, into = c("condition", "study_time"), sep = "_")
lb7

```

---
Then we can just make a normal boxplot.
```{r}
ggplot(lb7, aes(x = study_time, y = total, fill = condition)) +
  geom_boxplot()
```

---
## Two-way Within Subjects ANOVA
This is a combination of last weeks code with our between-subjects two-way code. We just specify our factors, interaction, and make sure to specify both in `(Error(ID/...))`.
```{r}
needs(afex)
m1 <- afex::aov_car(total ~ condition + study_time + condition:study_time + Error(ID/condition*study_time),
                           data=lb7, include_aov = F)
```

---

```{r}
summary(m1)
```

---
## Plot Marginal Means

Here I am creating a new summary data frame. After grouping by `condition` and `study_time`, I just make `total` equal to the mean of those groups. I am also reordering `study_time` so it renders in ascending order for the plot.

```{r}

means <- lb7 %>% 
  group_by(condition, study_time) %>% 
  summarise(total = mean(total)) %>% 
  mutate(study_time = factor(study_time, c("30", "60", "180")))
         
```

---

```{r}
ggplot(means, aes(x = study_time, y = total, group = condition)) +
  geom_point() +
  geom_line(aes(color = condition))
```

---

```{r}
ggplot(means, aes(x = condition, y = total, group = study_time)) +
  geom_point() +
  geom_line(aes(color = study_time))

```

---

## Pairwise Comparisons
Pairwise comparsions are similar to how we did contrasts the last few weeks. We run `emmeans` on our model using our formula (`~condition*study_time`) and then run pairs on that object.

```{r}
em <- emmeans(m1, ~condition*study_time)
pairs(em)
```

---

Again, `pairs` does not automatically generate confidence intervals. If you needed them, you can wrap it with `confint` or run it on your pairs object.

```{r}
confint(pairs(em))
```
---
## Alternative Pairwise Rendering

Remember, if you want to view pairwise comparisons in a different (more organized) format, you can use the following alternative specification.

```{r}

means_v2 <- emmeans(m1, "study_time", by = "condition")
pairs(means_v2)
```

---
```{r}
means_v3 <- emmeans(m1, "condition", by = "study_time")
pairs(means_v3)

```
---

## Contrasts
Contrasts should be pretty straightfoward and similar to previous weeks. Start by inspecting the order of your factors so you correctly specify the coding.

```{r}
em
```

```{r}
contrasts <- list(
  hyp1 = c(0, 1, -1, 0, -1 , 1)
)

```

---

```{r}
contrast(em, contrasts)
```

```{r}
effectsize::t_to_eta2(
  t = 4.773,
  df_error = 7
)

```
