---
title: "EDUC 640"
subtitle: "Three-Way ANOVA"
author: Chris Ives
output: 
  xaringan::moon_reader:
    css: ["default", "new.css"]
    lib_dir: libs
    nature:
      navigation:
        scroll: false
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
needs(tidyverse, xaringan, rio, here, misty, emmeans, rstatix, afex)
lb9 <- read.sav(here("data/Lab9_Potatoes.sav"), use.value.labels = T)
```

# Contents

* Expanding Boxplots
* Three-Way ANOVA
* Using afex
---

## Exploring the Design

Here we can use the `table` function to create three-way crosstabs.

```{r}
table(lb9$Size, lb9$Area, lb9$Temp)
```
---

## Plots

To create boxplots across three IVs, we will assign `Area` to the x-axis and facet `Temp` and `Size` using `facet_grid`. The code is a little more verbose than using `facet_wrap`.

```{r}
plot <- ggplot(lb9, aes(x = Area, y = Flavor)) +
  geom_boxplot() +
  facet_grid(rows = vars(Temp), cols = vars(Size))

```
---
```{r, echo = F}
plot
```

---
## Levene's Test

```{r}
car::leveneTest(Flavor ~ Area*Temp*Size, data = lb9, center = "mean")
```

---
## Three-Way Anova
Since we don't have any within-subjects effects, we can go back to using `rstatix::anova_test`. Writing `Area*Temp*Size` is a short hand for writing out all main effects and interactions.
```{r}
m_aov <- anova_test(Flavor ~ Area*Temp*Size, 
                    data = lb9, 
                    effect.size = "pes", 
                    type = 3, 
                    detailed = T)
```

---

```{r}
m_aov
```

---
## Estimated Means

To get `emmeans` to work, we need to run it on a `lm` object that is equivalent to our `anova_test`. The process from here is the same as we did in Week 5 if you wanted to do contrasts.

```{r}
m1 <- lm(Flavor ~ Area*Temp*Size, data=lb9)
emmeans(m1, ~Area*Temp*Size)
```

---
## Plotting Means
```{r}
means <- lb9 %>% 
  group_by(Area, Temp, Size) %>% 
  summarise(Flavor = mean(Flavor))

means_plot <- ggplot(means, aes(x = Area, y = Flavor, color = Temp)) +
  geom_point() + 
  geom_line(aes(group = Temp)) +
  facet_wrap(~Size)
```

---
```{r}
means_plot
```

---
## What about afex?

`afex::aov_car` is what we have used the last few weeks for our within-subjects and mixed ANOVAs. While it is convenient in that it works with `emmeans` it requires an ID variable (which we don't have for this dataset). I'll demonstrate the process if you wanted to continue using it.

---
## Adding an ID variable
Since every case is independent, I can just assign row numbers as the ID.
```{r}
lb9 <- lb9 %>% 
  mutate(id = row_number())

```
---

## Running afex
Notice if we only have between-subjects IVs, we just put the `id` term in `Error()`. The other thing we add is to change the `anova_table` to display partial eta instead of generalized eta.
```{r}
m_afex <- afex::aov_car(Flavor ~ Area*Temp*Size + Error(id), 
                        data = lb9, 
                        anova_table = list(es = "pes"))
m_afex

```

---

Then you can run `emmeans` directly on your afex object.
```{r}
emmeans(m_afex, ~Area*Temp*Size)
```

}